\Huge{Vipassana for Hackers} \\ \large{Paper Two: The Brain} \\ \large\textit{Version 0.2}

### 2019-10-03 paper-two/paper-two.tex
round two edits to paper two, done

@@

The CEN is activated when high-level cognitive tasks or external goal-oriented tasks
are being performed. These tasks or executive functions are cognitive processes involved in
cognitive control of behavior. This depends on three types of brain functions : working memory,
mental flexibility and self-control.
The regions of the brain involved in these executive functions are
Dorsolateral Prefrontal Cortex (DL-PFC), Orbito Frontal Cortex (OFC),
and the Posterior Parietal Cortex (PPC).

\begin{figure}[H]
  \centering
  \includegraphics[width=7.5cm]{images/top-sn.jpg}
  \caption{The Salience Network consists of the Anterior Cingulate Cortex and the
    Anterior Insular Cortex. It helps switch between the Default Mode Network and the
  Central Executive Network.}
  \label{fig:top-sn}
\end{figure}

@@

\subsubsection{Posterior Parietal Cortex}
The PPC is a region of the parietal cortex that is posterior to the primary somatosensory
cortex.
The posterior parietal cortex receives input from auditory, visual, and somatosensory systems
as well as a variety of other brain regions, and it integrates this input to facilitate
the execution of functions that require diverse information (also known as higher-order functions).
It in turn activates the DL-PFC and motor cortex and is involved in the performance of
attention related tasks as well as higher-order motor tasks such as grasping and catching. \cite{parietallobesa,parietallobes}

\subsection {Sidenote: Somatosensory Cortex}
The primary somatosensory cortex is located in a ridge of cortex found within the parietal lobe.
It is responsible for processing somatic sensations. These sensations arise from receptors
positioned throughout the body. They are responsible for detecting touch, for proprioception
(i.e. the position of the body in space), nociception (i.e. pain), and temperature. \cite{somato}
When such receptors detect one of these sensations,  the information is sent to the thalamus and
then to the primary somatosensory cortex.
The sensory homunculus which is a part of the somatosensory cortex is a cortical representation
of the body based on the degree of sensory innervation. \cite{sensoryhom}

\todo {include images of the homunculus here}

\todo{Summarize CEN? Relevance to meditation?}
So, the Central Executive Network is involved in goal-directed behavior and processes
related to attention. These goals are usually external and attention is also directed externally(whether
voluntary or not). When you meditate, you direct your attention voluntarily
inside, to the same sensory cues that are directing the PPC.

@@

The SN is an intrinsically connected large-scale network anchored in the Anterior
Insular Cortex (AIC) and Dorsal Anterior Cingulate Cortex (ACC). Both regions have reached a
high degree of specialization in the great apes. It is the collection of the regions
in the brain that help decide which stimuli deserves our attention. It acts as a
switch between the internally-directed DMN and the externally-directed Central
Executive Network. \cite{saliencenetwork}

@@

The Insula is a part of the Cerebral Cortex located deep within the Sulcus, the
fissure separating the four lobes of the brain. The AIC physically projects itself
into the Amygdala. It is involved in multimodal sensory processing such as
audio-visual integration tasks, interoceptive awareness (so its activity is directly
related to an individual's sense of internal body states), empathy and conscious
awareness. \cite{aicemotion}

@@

The salience network is directly involved in switching between the DMN and CN.
When you meditate, you cultivate the cognitive processes related to attention.
When your mind wanders, you notice it and bring it back.
The salience network should be directly involved in this process of assigning attention.
Meditation should strengthen the salience network and its ability to switch from
the default.

@@

I applied for my first 10-day Vipassana course after I quit my job. From the time I
was accepted into the course until the first day of the course I was very nervous
about what it might entail. I had never done anything like this (meditation) before
and I was quite happy with the idea that meditation, with its spiritual connotations
and religious mumbo-jumbo, was a waste of my time. In my mind, a silent meditation retreat was
for people who had time to waste --- and I was not one of them. Yet friends had convinced me that I
should give Vipassana a try and in the interval between workplaces I had
time to experiment with a course.

The following are my observations and speculations about what could be happening
in the brain during a 10 course.

I went for my first course in Chennai, my home town. When I arrived, I had to fill
all the application forms all over again (despite applying online), keep all my
luggage in a locker, and go to my room. I had to share the room with one other
person. I felt quite out of my element and I was certain everyone could see in me
that I was the odd one out, the one that was not supposed to be there. Because I felt
so out of place, I was too scared to actually make conversation with anyone before
silence was enforced.

@@

And so it was. With my brain filled with thoughts about myself, about the people
around me, I started this meditation business. For the first 3.5 days you are asked
to focus your attention on the area below your nostrils and above your upper
lip. That's all, these were really simple instructions, but so hard to keep up with.
Within the first two days, I had made up elaborate stories about my
fellow meditators, all of them superheroes, tirelessly working to save humanity
together. My default mode network was in overdrive.

Focusing our attention below the nostrils starts with observing your breath coming in
and going out. As the teacher mentions in the late evening discourses, the reason for
starting with your breath is because this the only activity of the body which is both
conscious and unconscious --- it acts as a bridge to the unconscious mind and the
involuntary processes of the body.

During the meditation hours I was trying hard. If I was to examine this practice
empirically, I had to give it an honest shot. I focused on the small patch beneath my
nostrils above my upper lip. To begin with, while trying to observe your breath and
your mind wanders, that is the default mode network working. As you realize your mind
has wandered and bring your attention back to your breath, your salience network is
activated. This happens slowly at first, with your mind wandering for many, many
minutes before you realize you are not with your breath. But within the 3.5 days, the
SN learns quite quickly to bring attention back to the breath. After a few days, this
refocusing almost happens by habit, almost automatically.

As I gave in to this activity, tried to focus more intently, tried to sit still for
longer, there were longer periods of awareness on that patch of skin and, with them,
stranger sensations arising and passing within that physical area. By the third day,
it felt like entire ecosystems were writhing and flopping and crashing, all of them
very alive in the area below my nostrils above my upper lip.

There have been studies that have shown that once you bring your
attention back to breath, the neural structures involved in the control
of the Autonomic Nervous System and attention start firing more actively
(in this case, the DL-PFC --- a part of the CEN and the Hippocampus a part of the DMN
and ACC --- a part of the SN). There are also global dampening changes seen in the brain,
particularly within the DMN and the Amygdala, which is involved in
the flight-or-fight (stress) response. These changes have been
termed the ``relaxation response'', as they are antagonistic to the stress
response. This relaxation response can be thought of as a gateway to altered states
of mind. \cite{relaxationresponse}

There has also been research in mice, showing a cluster of nerves called the
pre-Bötzinger Complex (preBötC), found in the brain stem of most mammals, which fires
with every breath taken. This breathing pacemaker seems to work not only for regular
conscious and unconscious breathing but for all kinds of breathing --- such as yawns,
sighs and gasps. The preBötc also appears to play a role in calming and
arousal. \cite{prebotcgeneration} It would be interesting to see the activation in
these neuron clusters during the first three days of a Vipassana course, even among
novice meditators.

@@

After lunch on the fourth day the instructions for Vipassana are given. During
Vipassana, you transfer your attention from the patch below your nostril to the top
of your head. From there, for two hours, the instructions are given to slowly move
your attention throughout the entire body: ``from the top of the head to the tips of
the toes'', then repeat. The two criteria for observation of bodily sensations are
focused attention (making use of the narrow focus practiced during the Anapana
period) and open monitoring (observing the sensations objectively).

The Salience Network (ACC and AIC) and the Central Executive Network (DL-PFC, OFC,
and PPC) which are involved in the control of attention start firing rapidly. Usually
when a person is engaged in an external task, the OFC and PPC are receiving signals
from all the sensory networks (Visual, Audio, Somatosensory, etc.) and, in response,
continuously send signals to the different motor cortices, Amygdala, and Hypothalamus to
respond by performing a task, feeling an emotion, etc.

But Vipassana is an internal task. The Salience Network is activated by focused
attention, which in turn activates the Central Executive Network. Rather than an
external task, as the CEN is accustomed to, the conscious instruction is to observe
bodily sensation --- and do nothing. So as you sit in silence with your eyes closed,
not moving and scanning your body with your attention, you are doing what the CEN has
always done except attempting not to react.

This is hard at first, with the body involuntarily reacting by jerking and writhing
in involuntary response to the act of observation. Initially, your awareness is
itself jerky and it is hard to observe sensations consciously. There is also the
pain. This eclipses all other sensations. In my case, it was difficult to maintain
awareness of any other sensation with the pain that was emanating from my legs and
back. The OFC is involved in assigning emotion to sensation \cite{ofcemotion} and the
ACC plays a vital role in perception of sensation. These two parts of the brain are
usually functionally connected. In the beginning, as I sat trying to observe
sensations and was only aware of the pain, the ACC was firing rapidly and the
OFC then assigned these signals to the Limbic System which
put me in a less than ideal emotional state.

But as your awareness gets smoother and observations get more objective, this
functional connectivity between the ACC and the OFC slowly starts changing. As you
continue to pay attention to the different sensations and you attempt to avoid assigning
emotions to them. Without assigning an emotional response, you can observe with a
non-judgmental attitude, particularly toward unpleasant stimuli. The OFC still fires
but the Limbic System is quiesced and, in turn, you react less. It is as if both your
conscious and unconscious awarenesses are slowly aligning until you experience some
occasions when it feels as if there is nothing but detached awareness moving
through your body (or some parts of your body). Senses are not dulled, however. In
addition to a feeling of emotional detachment, the Insula and the Somatosensory
Cortex, which are involved in sensory awareness, were firing \textbf{more} actively,
not less.

Perhaps the strangest experience during a Vipassana course is that of shifts in your
sense of self. The ``self'', as a subjective agent in space and time changes, if at
first only very slightly. As you close your eyes and meditate, it may feel as though
the body has shifted in space or rotated in different directions. During these times,
the vmPFC and PCC become less active. As the meditations get deeper the DMN
occasionally quiets completely (perhaps only for a short duration) while the Insula
fires more rapidly, leading to alternative networks of consciousness arising. It is
as if losing the influence of one's own narrative leads to new insights that are
usually kept from consciousness. It has been found that experiencing this, even for a
few seconds, can lead to lasting changes long after the activity of meditation itself
has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
similar to being on certain hallucinogenic drugs or even, in some cases, to ecstatic
seizures where the abnormal activity of the Anterior Insular Cortex leads to
heightened self-awareness, feelings of bliss, and a lack of ambiguity. \cite{cortexbliss}

@@

Vipassana causes global dampening changes in the DMN, causes certain parts of
the SN and CEN to fire more rapidly. That is, reacting parts of the network and
portions of the network which are aware of bodily sensation, respectively.
This changes a person's behavior (in the short term, at least) and these
changes can often be seen as soon as someone leaves the course on the tenth day.

But repeated practice is essential for any sustained neuroplasticity since rewriting
many years of habit formation requires more than a ten day course. However, the ten
day course is well-designed and will give someone experimenting with meditation the
minimum observational change necessary to instill an interest, to keep practicing.

There have been studies looking at both the structural changes in the brain and
functional changes in brain activity, especially to the DMN and the SN. But such
studies are usually only done on meditators who have clocked thousands of hours of
practice, which diverges from the purpose of this paper. \cite{tibetanmonks,
  experiencemed, experiencemeda}

How is Vipassana different from simply concentrating on a task, as hard as one can?
When concentrating on any task, your DMN is quieted, the SN and CEN are firing, and you
enter into a state of flow. But flow or no flow, the reacting part of the brain is
also working. And it is working hard. The goal of Vipassana is not simply to
concentrate or to enter a state of flow, but to break old reactionary habit patterns
altogether.

As I left the Vipassana Center, I was really intrigued as to what exactly had
occurred over those ten days. What was it that had changed in my brain? I had to
know. I very quickly noticed, however, that as I left the centre (or even earlier, as
soon as talking was allowed), those very changes started slipping away. But with each
hour-long meditation at home, a tiny bit sticks just a little more. And I get closer
to my answer.

>>>>

Where possible in the following sections these networks are described in terms of
function. The description of the DMN is also organized by function but the subsystems
of the CEN and SN offer too many potential functional categorizations and are instead
organized by anatomy.

@@

The CEN is activated when high-level cognitive tasks or external goal-oriented tasks
are being performed. These tasks or executive functions are cognitive processes involved in
cognitive control of behavior. This depends on three types of brain functions:
working memory, mental flexibility, and self-control. The regions of the brain
involved in these executive functions are the Dorsolateral Prefrontal Cortex
(DL-PFC), Orbito Frontal Cortex (OFC), and the Posterior Parietal Cortex (PPC).

\begin{figure}[H]
  \centering
  \includegraphics[width=5cm]{images/top-cen.jpg}
  \caption{The Central Executive Network.}
  \label{fig:top-cen}
\end{figure}

@@

\subsubsection{Posterior Parietal Cortex}
The PPC is a region of the Parietal Cortex, physically located behind the primary
Somatosensory Cortex. The PPC receives input from auditory, visual, and somatosensory systems
as well as a variety of other brain regions \todo{What other brain regions provide
  input to the PPC?}, and it integrates this input to facilitate
the execution of functions that require diverse information (also known as higher-order functions).
In turn, the PPC activates the DL-PFC and motor cortex and is involved in the performance of
attention related tasks as well as higher-order motor tasks such as grasping and catching. \cite{parietallobesa,parietallobes}

\todo{PSC section mentions Parietal Lobe but it hasn't been mentioned earlier, nor defined?}

\subsubsection{Sidenote: Somatosensory Cortex}
The Primary Somatosensory Cortex is located in a ridge of cortex found within the Parietal Lobe.
It is responsible for processing somatic sensations. These sensations arise from receptors
positioned throughout the body. They are responsible for detecting touch, for proprioception
(the position of the body in space), nociception (pain), and temperature. \cite{somato}
When such receptors detect one of these sensations, the information is sent first to
the Thalamus and then to the Primary Somatosensory Cortex.

The sensory homunculus which is a part of the somatosensory cortex is a cortical representation
of the body based on the degree of sensory innervation. \cite{sensoryhom}

\todo{Include side image of PSC here.}

\todo{Include images of the homunculus here}

The Central Executive Network is involved in goal-directed behavior and processes
related to attention. These goals are usually external and attention is also directed
externally (regardless of whether attention is voluntary or not). When you meditate,
attention is directed both internally and voluntarily, to the same sensory cues that
are normally directing the PPC. \todo{What is interesting about internal/voluntary
  attention in CEN?}

@@

The SN is an intrinsically connected large-scale network anchored in the Anterior
Insular Cortex (AIC) and Dorsal Anterior Cingulate Cortex (ACC). Both regions have reached a
high degree of specialization in the great apes. It is the collection of the regions
in the brain that help decide which stimuli deserves our attention. It acts as a
switch between the internally-directed DMN and the externally-directed CEN. \cite{saliencenetwork}

@@

The AIC is a part of the Cerebral Cortex located deep within the Sulcus, the
fissure separating the four lobes of the brain. The AIC physically projects itself
into the Amygdala. It is involved in multimodal sensory processing such as
audio-visual integration tasks, interoceptive awareness (so its activity is directly
related to an individual's sense of internal body states), empathy, and conscious
awareness. \cite{aicemotion}

@@

The SN is directly involved in switching between the DMN and CEN. When you meditate,
you cultivate the cognitive processes related to attention.
When your mind wanders, you notice it and bring it back.
The SN is directly involved in this process of assigning attention, since it is
responsible for the prominence of any object of attention.
Meditation, or the act of returning to an object of meditation, can therefore be
thought of as an exercise in strengthening the SN and its ability to switch from
the default objects of attention, over time.

@@

I applied for my first 10-day Vipassana course after I quit my job. From the time I
was accepted into the course until the first day of the course I was very nervous
about what it might entail. I had never done anything like meditation before
and I was quite happy with the idea that meditation, with its spiritual connotations
and religious mumbo-jumbo, was a waste of my time. In my mind, a silent meditation retreat was
for people who had time to waste --- and I was not one of them. Yet friends had convinced me that I
should give Vipassana a try and in the interval between workplaces I had
time to experiment with a course.

The following are my observations and speculations about what could be happening
in the brain during a 10-day course.

I went for my first course in Chennai, my home town. When I arrived, I had to fill
the application forms all over again (despite applying online), keep all my
luggage in a locker, and go to my room. I had to share the room with one other
student. I felt quite out of my element and I was certain everyone else could see
that I was the odd one out. In my head, I was not supposed to be there. Because I felt
so out of place, I was too scared to actually make conversation with anyone before
silence was enforced.

@@

And so it was. With my brain filled with thoughts about myself, about the people
around me, I started this meditation business. For the first 3.5 days students are asked
to focus their attention on the area below their nostrils and above their upper
lip. That's all. These are undoubtedly very simple instructions but in practice I
found them very difficult. Maintaining my focus was much harder than I had
anticipated. Within the first two days, I had made up elaborate stories about my
fellow meditation students. In these fantasies all of the other students were
superheroes, tirelessly working to save humanity together. My Default Mode Network
was in overdrive.

Focusing our attention below the nostrils starts with observing breath --- coming in
and going out. As the teacher mentions in the late evening discourses, the reason for
starting with your breath is because this the only activity of the body which is both
conscious and unconscious --- it acts as a bridge to the unconscious mind and the
involuntary processes of the body.

\todo{Elaborate on breath/bridge and how it relates to brain function?}

During the meditation hours I was trying hard. I had decided that if I was to examine this practice
empirically, I had to give it an honest shot. I focused on the small patch beneath my
nostrils above my upper lip. Initially, while trying to observe one's breathing, the
mind will wander --- not just to fantasies of superheroes but to almost any object or
daydream imaginable --- as the mind wanders the Default Mode Network is working. With
the realization that the mind has wandered, and the consequent return of attention
back to the breath, the Salience Network is
activated. This happens slowly at first. One's mind wanders away for many, many
minutes before realizing that the object of attention, the breath, has been lost. But
at some point within the 3.5 days, the SN learns to bring attention back to the breath. After a few days, this
refocusing almost happens by habit, almost automatically.

As I gave in to this activity, tried to focus more intently, and tried to sit still for
longer continuous stretches of time, there were longer periods of awareness on that
patch of skin. With these longer periods of awareness came
stranger sensations arising and passing within that physical area. By the third day,
it felt like entire ecosystems were writhing and flopping and crashing, all of them
very alive in the area below my nostrils above my upper lip.

\todo{Neurological / physiological basis for these strange sensations?}

There have been studies that have shown that once subjects bring their
attention back to breath, the neural structures involved in the control
of the Autonomic Nervous System and attention start firing more actively
(in this case, the DL-PFC --- a part of the CEN, the Hippocampus --- a part of the
DMN, and the ACC --- a part of the SN). During Anapana there are also global
dampening changes seen in the brain, particularly within the DMN and the Amygdala, which is involved in
the flight-or-fight (stress) response. These changes have been
termed the ``relaxation response'', as they are antagonistic to the stress
response. This relaxation response can be thought of as a gateway to altered states
of mind. \cite{relaxationresponse}

\todo{Relate ``gateway to altered states of mind'' to the Anapana ``bridge''?}

There has also been research in mice, showing a cluster of nerves called the
pre-Bötzinger Complex (preBötC), found in the brain stem of most mammals, which fires
with every breath taken. This breathing pacemaker seems to work not only for regular
conscious and unconscious breathing but for all kinds of breathing --- such as yawns,
sighs and gasps. The preBötC also appears to play a role in calming and
arousal. \cite{prebotcgeneration} It would be interesting to see the activation in
these neuron clusters during the first three days of a Vipassana course, even among
novice meditators.

@@

After lunch on the fourth day the instructions for Vipassana are given. During
Vipassana, students transfer their attention from the patch below their nostrils to the top
of the head. From there, for two hours, instructions are given to slowly move
one's attention throughout the entire body: ``from the top of the head to the tips of
the toes.'' This movement of attention throughout the body is then repeated, over and
over. The two objectives while observing bodily sensations are focused attention (making
use of the narrow focus practiced during the Anapana period) and open monitoring
(observing the sensations objectively).

As this process progresses, the Salience Network (ACC and AIC) and the Central Executive Network (DL-PFC, OFC,
and PPC), which are involved in the control of attention, fire more actively. Usually
when a person is engaged in an external task, the OFC and PPC receive signals
from all the sensory networks (Visual, Audio, Somatosensory, etc.) and, in response,
continuously send signals to the different motor cortices, Amygdala, and Hypothalamus to
respond by performing a task, feeling an emotion, etc.

But Vipassana is an internal task. The Salience Network is activated by focused
attention, which in turn activates the Central Executive Network. Rather than an
external task, as the CEN is accustomed to, the conscious instruction is to observe
bodily sensation --- and do nothing. So as you sit in silence with your eyes closed,
not moving your body physically but instead moving throughout your body with your
attention, the CEN is doing what it has always done. However, the outcomes of
activity in the CEN is different during Vipassana. Instead of focusing on
external sensory input the sensory input is internal (bodily sensations) and instead
of reacting to input by performing tasks or feeling emotions the goal is to not
react, to simply observe as objectively as possible.

This is hard at first, with the body involuntarily reacting by jerking and writhing
in involuntary response to the very act of observation. Initially, even awareness
itself has a jerky quality and it is hard to observe sensations consciously. There is also the
pain. This eclipses all other sensations. In my case, it was difficult to maintain
awareness of any other sensation with the pain that was emanating from my legs and
back. The OFC is involved in assigning emotion to sensation \cite{ofcemotion} and the
ACC plays a vital role in perception of sensation. These two parts of the brain are
usually functionally connected. In the beginning, as I sat trying to observe
sensations and was only aware of the pain, the ACC was firing rapidly and the
OFC then assigned these signals to the Limbic System, which
put me in a less than ideal emotional state.

Over the final seven days of the course, as awareness gets smoother and observations get more objective, this
functional connectivity between the ACC and the OFC slowly starts changing. While
continuing to pay attention to constantly changing bodily sensations the goal remains: avoid
assigning evaluations (and therefore, emotions) to them. Without assigning an
emotional response, one learns to observe with a
non-judgmental attitude, particularly toward unpleasant stimuli. The OFC still fires
but the Limbic System is quiesced and, in turn, reactions become smaller and less
frequent. This snowball effect almost mirrors the spiral of brain behaviours
exhibited early in the course. Early in the course, the CEN's normal mode of
operation is in play, which is to experience unpleasant stimuli and react (desiring to rid the body of
them), which in turn causes the unpleasant stimuli to appear all the more
unpleasant. Once a measure of objectivity is reached, however, the brain's function
snowballs in the other direction. It is as if both conscious and unconscious awarenesses are slowly aligning
until there are occasions of experience when it feels as if there is nothing but detached awareness moving
through the body (or some parts of the body). This is not at all akin to the
detachment of sleep or intoxication, however, and the senses are not dulled. In
addition to a feeling of emotional detachment, the Insula and the Somatosensory
Cortex, which are involved in sensory awareness, were \textbf{more} active, not
less. As the ACC, OFC, and Limbic System are calmed and attention is narrowed
further, the sensations become progressively more intense.

Perhaps the strangest experience during a Vipassana course is that of shifts in the
sense of self. The ``self'', as a subjective agent in space and time, changes. At
first, this change is only very slight. As a student closes her eyes and begins to
meditate, it may feel as though
the body has shifted in space or rotated in different directions. During these times,
the vmPFC and PCC become less active. \todo{So? What does this mean?} As the meditations get deeper the DMN
occasionally quiets completely (perhaps only for a short duration) while the Insula
fires more rapidly, leading to alternative networks of consciousness
arising. \todo{What does this mean? What is an example of ``alternative networks of consciousness''?} It is
as if losing the influence of one's own narrative leads to new insights that are
usually kept from consciousness. It has been found that experiencing this, even for a
few seconds, can lead to lasting changes long after the activity of meditation itself
has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
similar to being on certain hallucinogenic drugs or even, in some cases, to ecstatic
seizures where the abnormal activity of the Anterior Insular Cortex leads to
heightened self-awareness, feelings of bliss, and a lack of ambiguity. \cite{cortexbliss}

@@

Vipassana causes global dampening changes in the DMN, causes certain parts of
the SN and CEN to activate. That is, reacting parts of the network and
portions of the network which are aware of bodily sensation, respectively.
This changes a person's behavior and these changes can often be seen as soon as
someone leaves the course on the tenth day.

But repeated practice is essential for any sustained neuroplasticity since rewriting
many years of habit formation requires more than a ten day course. However, the ten
day course is well-designed \todo{How is the course well designed?} and will give someone experimenting with meditation the
minimum observational change necessary to instill an interest, to keep practicing.

There have been studies looking at both the structural changes in the brain and
functional changes in brain activity, especially to the DMN and the SN. But such
studies are often only done on meditators who have clocked thousands of hours of
practice, which diverges from the purpose of this paper. \cite{tibetanmonks,
  experiencemed, experiencemeda}

How is Vipassana different from simply concentrating on a task, as hard as one can?
When concentrating on any task, your DMN is quieted, the SN and CEN are firing, and you
enter into a state of flow. But flow or no flow, the reacting part of the brain is
also working. The mind may learn to focus intensely but the habit patterns of the
mind are unlikely to change. The goal of Vipassana is not simply to concentrate or to
enter a state of flow, but to break old reactionary habit patterns altogether.

As I left the Vipassana Center, I was really intrigued as to what exactly had
occurred over those ten days. What was it that had changed in my brain? I had to
know. I very quickly noticed, however, that as I left the centre, those very changes
started slipping away. But with each hour-long meditation at home, a tiny bit sticks
just a little more. And I get closer to my answer.
